---
title: "Assumptions & Diagnostics"
output: ioslides_presentation
---


## BLUE
- Best linear unbiased estimate of beta  
<br> 
Properties: Unbiased, Efficient (se minimum), Consistent (N decreases se)

## 6 assumptions
1. No measurement error
2. Correctly specified form
3. No omitted variables
4. Homoscedasticity 
5. Independence among the residuals
6. Normally distributed residuals

## What happens if we violate assumptions?
1. Biased regression coefficients
2. Biased standard errors

## 6 assumptions

|Violated Regression Assumption |	Coefficients |	Standard Errors|
|-------------------------------|--------------|-----------------|
|1. Measured without error     	|	 Biased 		 |	  Biased       |
|2. Correctly specified form 		|	 Biased 		 |	  Biased       |
|3. Correctly specified model		|	 Biased			 |	  Biased       | 
|4. Homoscedasticity 						|				       |    Biased       |
|5. Independent Errors 				 	|				       |    Biased       |
|6. Normality of the Errors 		|							 |    Biased       |


## How do we detect violations? 
|       Assumption              |       	detect                    |
|-------------------------------|--------------------------------|
|1. Measured without error	    |	  Reliability                  |
|2. Correctly specified form 		|	  Residuals against predicted  |
|3. Correctly specified model		|	  Theory                       | 
|4. Homoscedasticity 						|		Residuals against predicted  |
|5. Independent Errors 				 	|		Design                       |
|6. Normality of the Errors     |   q-q plot or distribution     |

##
load data
```{r, messages = FALSE, warning= FALSE}

library(readr)
a_data <- read.csv("diagnostics.csv")
library(broom)
model.1 <- lm(Anxiety ~ Stress + Support, a_data)
a_data<- augment(model.1)
a_data
```
## residuals are very important

```{r, echo = FALSE, warning=FALSE}

model.0 <- lm(Anxiety ~ Support , a_data)
a_data.0<- augment(model.0)

library(ggplot2)
ggplot(a_data.0, aes(x = Support, y = Anxiety)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightgrey") +  # Plot regression slope
  geom_segment(aes(xend = Support, yend = .fitted), alpha = .2) +  # alpha to fade lines
  geom_point() +
  geom_point(aes(y = .fitted), shape = 1)
```

## with stress in the model
```{r, echo = FALSE}

ggplot(a_data, aes(x = Support, y = Anxiety)) +
  geom_segment(aes(xend = Support, yend = .fitted), alpha = .2) +  # alpha to fade lines
  geom_point() +
  geom_point(aes(y = .fitted), shape = 1)
```



## Assumption 1: No measurement error in our independent variables



## Assumption 2: Correctly specified form

```{r, echo = FALSE}
library(ggplot2)

ggplot(data = a_data, aes(x=.fitted, y=.resid)) + 
  geom_hline(yintercept = 0) +
  geom_point() + geom_smooth( se = FALSE)
```



## Assumption 3: Correctly specified model





## Assumption 4: Homoscedasticity

```{r, echo = FALSE}
ggplot(data = a_data, aes(x=.fitted, y=.resid)) + geom_point() + 
    geom_hline(yintercept = 0) +
  geom_smooth(se = FALSE)

```



## Assumption 5: Independence among the errors

```{r, echo = FALSE}
a_data$ID <- c(1:118)
ggplot(data = a_data, aes(x=ID, y = .resid)) + geom_point() +  geom_hline(yintercept = 0) 

```


## Assumption 6: Normality of the errors
```{r, echo = FALSE}
ggplot(data = a_data, aes(x= .resid)) + geom_density() +   xlim(-10, 10)
```



##
```{r, echo = FALSE}
ggplot(model.1) +
  stat_qq(aes(sample = .stdresid)) +
  geom_abline()
```


## car package also works 
```{r}
library(car)
qqPlot(model.1)
#base plot function too
#plot(model.1, which = 2)
```



## How do we fix violations? 

|       Assumption              |       	Fix                      |
|-------------------------------|----------------------------------|
|1. Measured without error	    |	Factor scores, SEM, more data    |
|2. Correctly specified form 		|	Different model                  |
|3. Correctly specified model		|	 ¯`\_`(ツ)`_`/¯                  |
|4. Homoscedasticity 						|	Bootstraps, WLS,transformations  |
|5. Independent Errors 				 	| Use differ method                |
|6. Normality of the Errors     | Additional IVs, differnt form    |

## Robustness? 

- yeah, mostly
- especially with normally of errors
- note: we do not need to have normally distributed IVs or DVs


## Data screeing

- Start with descriptive stats and plots   
- Look for outliers, missing cases, etc.  
- Test assumptions (residual plots)  

## why screen your data? 

## Outliers
- Broadly defined as atypical or highly influential data point(s) 
- Due to contamination (e.g. recording error) or accurate observation of a rare case 
- Univariate vs. Multivariate 

## Outliers

1. Leverage   
    + How unusual is this case from the rest of the cases in terms of predictors?  
2. Distance   
    + How distant is the observed case from the predicted value?   
3. Influence   
    + How much the does regression coefficient change if case were removed? 
    
## Leverage

- Tells us how far observed values for a case are from mean values on the set of IVs (centroid)   
- Not dependent on Y values  
- High leverage cases have greater potential to influence regression results  
- Mahalanobis Distance for multivariate stats 

## Distance

- Distance From Prediction- how far is observed value from predicted value (i.e. residuals)   

- deleted residuals takes into consideration what would happen if case were deleted 



## Influence
- How regression equation would change if extreme case were removed 
- Influence = Leverage X Distance 
- DFFITS (difference in fit standardized)
- Cooks D (same but different metric to DFFITS)
- DFBETAS (coefficient change without outlier)



## cooks d
```{r, echo = FALSE}
ggplot(a_data, aes(x = ID, y = .cooksd)) +
geom_point() +
geom_text(aes(label = rownames(a_data)), vjust = -1)
```


## df beta
```{r, echo = FALSE}

a_data$dfbetastress <- dfbeta(model.1)[,"Stress"]

ggplot(data = a_data, aes(x = ID, y = dfbetastress)) + geom_point() +
geom_text(aes(label = rownames(a_data), vjust = -1))
```

## recommendations
- Analyze data with/without outliers and see how results change   
- If you throw out cases you must believe it is not representative of population of interest or have appropriate explanation 


