---
title: "Assumptions"
output: html_document
---


load data
```{r}
library(readr)
a_data <- read.csv("~/Box Sync/5067 Regression Spring/regression/Part 1 data.csv")
library(broom)
model.1 <- lm(Anxiety ~ Stress + Support, a_data)
a_data<- augment(model.1)
a_data
```

Assumption 1: No measurement error in our independent variables




Assumption 2: Correctly specified form

```{r}
library(ggplot2)

ggplot(data = a_data, aes(x=.fitted, y=.resid)) + geom_point() + geom_smooth(method='lm', se = FALSE)
```



Assumption 3: Correctly specified model





Assumption 4: Homoscedasticity, i.e. constant variance of the errors

```{r}
ggplot(data = a_data, aes(x=.fitted, y=.resid)) + geom_point() + geom_smooth(method='lm', se = FALSE)

ggplot(data = a_data, aes(x=Stress, y=.resid)) + geom_point() + geom_smooth(method='lm', se = FALSE)

ggplot(data = a_data, aes(x=Support, y=.resid)) + geom_point() + geom_smooth(method='lm', se = FALSE)
```



Assumption 5: Independence among the errors

```{r}
nrow(a_data)
a_data$ID <- c(1:118)
ggplot(data = a_data, aes(x=ID, y = .resid)) + geom_point()
```


Assumption 6: Normality of the errors
```{r}
ggplot(data = a_data, aes(x= .resid)) + geom_density() +   xlim(-10, 10)
```



```{r}
library(car)
qqPlot(model.1)
```


```{r}
plot(model.1)
```

Outliers

leverage, distance, and influence.

Influence
cooks d
```{r}
ggplot(a_data, aes(x = ID, y = .cooksd)) +
geom_point() +
geom_text(aes(label = rownames(a_data)), vjust = -1)
```


df beta
```{r}

a_data$dfbetastress <- dfbeta(model.1)[,"Stress"]

ggplot(data = a_data, aes(x = ID, y = dfbetastress)) + geom_point() +
geom_text(aes(label = rownames(a_data), vjust = -1))
```

